---
title: "Course Project for Statistical Inference Course on Coursera.org"
author: "Nan"
date: "25 Jan 2015"
output: html_document
---

#Overview
In this work, we investigate the __sample mean__ and __sample variance__ from 
the exponential distribution; and we compare their expectations from 
the simultation with the __population mean__ and __population variance__, 
respectively. 
To further check the variability of the __sample mean__, we look into the variance of the __sample mean__ and 
compare with the logical estimate from the Central Limit Theorem (CLT).

#Simulations
We choose the exponential distribution with the probability density function 
$p(x) = \lambda \exp^{-\lambda x}$. In all the simulations, $\lambda = 0.2$. 

In total, we run *num_simulations* simulations, in each of which there are 
*num_samples* samples.

```{r, simulation}
set.seed(0)
run_simulation <- function(num_simulations, num_samples, eval_option, lambda=0.2) {
    sample_means = NULL
    sample_variances = NULL
    for (sim_idx in 1:num_simulations) {
        sim_data = rexp(num_samples, rate = lambda)
        sample_means = c(sample_means, mean(sim_data))
        sample_variances = c(sample_variances, var(sim_data))
    }
    sample_means <- data.frame(mean=sample_means) 
    sample_vars <- data.frame(var=sample_variances)
    results <- NULL
    if(eval_option=="mean")
        results <- sample_means
    if(eval_option=="var")
        results <- sample_vars
    return(results)
}
```

```{r, multiplot_def, echo=FALSE}
library(ggplot2)
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
# Reference: http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
    require(grid)
    
    # Make a list from the ... arguments and plotlist
    plots <- c(list(...), plotlist)
    
    numPlots = length(plots)
    
    # If layout is NULL, then use 'cols' to determine layout
    if (is.null(layout)) {
        # Make the panel
        # ncol: Number of columns of plots
        # nrow: Number of rows needed, calculated from # of cols
        layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                         ncol = cols, nrow = ceiling(numPlots/cols))
    }
    
    if (numPlots==1) {
        print(plots[[1]])
        
    } else {
        # Set up the page
        grid.newpage()
        pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
        
        # Make each plot, in the correct location
        for (i in 1:numPlots) {
            # Get the i,j matrix positions of the regions that contain this subplot
            matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
            
            print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                            layout.pos.col = matchidx$col))
        }
    }
}


```

#Comparison between Sample Mean and Population Mean
The population mean $\mu$ is $\frac{1}{\lambda}$, which can be calculated 
analytically. According to the CLT, the expectation of the sample mean is equal
to $\mu$; the variance of the sample mean is $\frac{\sigma^2}{n}$, where 
$\sigma^2$ is the population variance and $n$ is the number of samples.

Here we show the distribution of the __sample mean__. Note the red line denotes the 
__population mean__. 

```{r, plot_mean, echo=FALSE, warning=FALSE, message=FALSE, fig.width=9, fig.height=3}
num_simulations <- 1000
sample_means_1 <- run_simulation(num_simulations, 10, "mean")
fig_mean_1 <- ggplot(sample_means_1, aes(x=mean)) + 
    geom_histogram(aes(y=..density..), binwidth=0.2) +
    geom_density(size=1, alpha=.2, fill="salmon") +
    geom_vline(xintercept=5, size=1, color="red") +
    xlim(0, 10) +
    ggtitle("n=10") 
sample_means_2 <- run_simulation(num_simulations, 20, "mean")
fig_mean_2 <- ggplot(sample_means_2, aes(x=mean)) + 
    geom_histogram(aes(y=..density..), binwidth=0.2) +
    geom_density(size=1, alpha=.2, fill="salmon") +
    geom_vline(xintercept=5, size=1, color="red") +
    xlim(0, 10) +
    ggtitle("n=20")
sample_means_3 <- run_simulation(num_simulations, 40, "mean")
fig_mean_3 <- ggplot(sample_means_3, aes(x=mean)) + 
    geom_histogram(aes(y=..density..), binwidth=0.2) +
    geom_density(size=1, alpha=.2, fill="salmon") +
    geom_vline(xintercept=5, size=1, color="red") +
    xlim(0, 10) +
    ggtitle("n=40")
multiplot(fig_mean_1, fig_mean_2, fig_mean_3, cols=3)
```

The comparison is further shown in the following table.

n | expectated sample mean | population mean
- | ---------------------- | ---------------
10| `r mean(sample_means_1$mean)`| 5                
20| `r mean(sample_means_2$mean)`| 5
40| `r mean(sample_means_3$mean)`| 5

> The results suggest that the expectation of the __sample mean__ is approaching 
to __population mean__ when the number of samples increases.

#Comparison between Sample Variance and Population variance
The popluation variance $\sigma^2$ is $\frac{1}{\lambda^2}$. 

We plot the distribution of the __sample variance__ in 
black. For comparison, the __population mean__ is plotted in red line.

```{r, plot_variance, echo=FALSE, warning=FALSE, message=FALSE, fig.width=9, fig.height=3}
sample_vars_1 <- run_simulation(num_simulations, 10, "var")
fig_var_1 <- ggplot(sample_vars_1, aes(x=var)) + 
    geom_histogram(aes(y=..density..), binwidth=1) +
    geom_density(size=1, alpha=.2, fill="salmon") +
    geom_vline(xintercept=25, size=1, color="red") +
    xlim(0,100)+
    ggtitle("n=10") 
sample_vars_2 <- run_simulation(num_simulations, 20, "var")
fig_var_2 <- ggplot(sample_vars_2, aes(x=var)) + 
    geom_histogram(aes(y=..density..), binwidth=1) +
    geom_density(size=1, alpha=.2, fill="salmon") +
    geom_vline(xintercept=25, size=1, color="red") +
    xlim(0,100)+
    ggtitle("n=20")
sample_vars_3 <- run_simulation(num_simulations, 40, "var")
fig_var_3 <- ggplot(sample_vars_3, aes(x=var)) + 
    geom_histogram(aes(y=..density..), binwidth=1) +
    geom_density(size=1, alpha=.2, fill="salmon") +
    geom_vline(xintercept=25, size=1, color="red") +
    xlim(0,100)+
    ggtitle("n=40")
multiplot(fig_var_1, fig_var_2, fig_var_3, cols=3)
```

The comparison is further shown in the following table.

n | expectated sample variance | population variance
- | ---------------------- | ---------------
10 | `r mean(sample_vars_1$var)` | 25               
20| `r mean(sample_vars_2$var)` | 25
40| `r mean(sample_vars_3$var)` | 25

> The results suggest that the expectation of the __sample variance__ is approaching
to __population variance__ when the number of samples increases. Moreover, the 
distributions are increasingly concerntrated. We also note that the distributions
are not normal distributions and are skewed.

# The Variance of Sample Mean
According to CLT, the variance of the __sample mean__ is $\frac{\sigma^2}{n}$.
We plot the distribution of the logical estimates of the variances, i.e.
$\frac{sample variance}{n}$.

```{r, plot_variance_mean, warning=FALSE, echo=FALSE, message=FALSE, fig.width=9, fig.height=3}
sample_vars_1$var <- sample_vars_1$var / 10
fig_var_1 <- ggplot(sample_vars_1, aes(x=var)) + 
    geom_histogram(aes(y=..density..), binwidth=0.2) +
    geom_density(size=1, alpha=.2, fill="salmon") +
    geom_vline(xintercept=25/10, size=1, color="red") +
    ggtitle("n=10") 
sample_vars_2$var <- sample_vars_2$var / 20
fig_var_2 <- ggplot(sample_vars_2, aes(x=var)) + 
    geom_histogram(aes(y=..density..), binwidth=0.2) +
    geom_density(size=1, alpha=.2, fill="salmon") +
    geom_vline(xintercept=25/20, size=1, color="red") +
    ggtitle("n=20")
sample_vars_3$var <- sample_vars_3$var / 40
fig_var_3 <- ggplot(sample_vars_3, aes(x=var)) + 
    geom_histogram(aes(y=..density..), binwidth=0.2) +
    geom_density(size=1, alpha=.2, fill="salmon") +
    geom_vline(xintercept=25/40, size=1, color="red") +
    ggtitle("n=40")
multiplot(fig_var_1, fig_var_2, fig_var_3, cols=3)
```

The comparison is further shown in the following table.

n | average of the logical estimates | variance of sample mean
- | ---------------------- | ---------------
10 | `r mean(sample_vars_1$var)` | 2.50               
20| `r mean(sample_vars_2$var)` | 1.25
50| `r mean(sample_vars_3$var)` | 0.625

> The results suggest that logical estimate of the __sample mean__ is a reasonable
estimate. The __sample mean__ become less variable, when the number of samples 
increases, because and the difference between 
this estimate and the estimand becomes increasingly
smaller. As the distribution of the __sample variance__,
these distributions are not normal distribution and are skewed.

#Conclusion
* We show that the distribution of __sample mean__ are approaching to a normal 
distribution when the number of samples increases. Meanwhile, 
    + the expectation of __sample mean__ is approaching to __population mean__;
    + the __sample mean__ become less variable. 
* The distribution of  __sample variance__ is not normal distributed and skewed. 
But their expectation is approaching to the __population variance__.
* These results are consistent with the CLT.

#Appendix
The function for plotting multiple figures in defined here.
```{r, multiplot, eval=FALSE}
library(ggplot2)
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
# Reference: http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
    require(grid)
    
    # Make a list from the ... arguments and plotlist
    plots <- c(list(...), plotlist)
    
    numPlots = length(plots)
    
    # If layout is NULL, then use 'cols' to determine layout
    if (is.null(layout)) {
        # Make the panel
        # ncol: Number of columns of plots
        # nrow: Number of rows needed, calculated from # of cols
        layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                         ncol = cols, nrow = ceiling(numPlots/cols))
    }
    
    if (numPlots==1) {
        print(plots[[1]])
        
    } else {
        # Set up the page
        grid.newpage()
        pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
        
        # Make each plot, in the correct location
        for (i in 1:numPlots) {
            # Get the i,j matrix positions of the regions that contain this subplot
            matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
            
            print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                            layout.pos.col = matchidx$col))
        }
    }
}


```